A Dependently Typed Multi-Stage Calculus\thanks{Supported by organization x.}

### 2019-06-02 main.tex
Fix section 2 with grammarly.

In \LTP, quote and unquote are written $\TB_\alpha M$ and $\TBL_\alpha M$, respectively.
The type of $\TB_\alpha M$ is $\TW_\alpha \tau$ when $M$ has $\tau$ type.
The type of $\TBL_\alpha M$ is $\tau$ when $M$ has $\TW_\alpha \tau$.
Please notice that $\TBL_\alpha M$ is well-typed, $M$ has a code type from the typing rule of \LTP.
In addition to normal $\beta$-reduction, there is an reduction rule for unquoted quoted code.
\begin{align*}
	\TBL_\alpha (\TB_\alpha M) \longrightarrow M
\end{align*}
It means when a quoted code fragment is unquoted, it returns back to the original code fragment.
This reduction is called $\longrightarrow_\Lambda$ in \LMD.
\red{この文はここに来るべきなのか? LTPの説明ではないが}

@@

A sequence of transition variables is a transition which is called a stage in \LMD.
In \LTP, all judgement has a transition.
Especially, terms without quoting exist at the empty transition which is represented by $\epsilon$.
For example, $(\lambda x:\I.x)\ (1+2)$ is at $\epsilon$ transition and 
$\TB_\alpha (\lambda x:\I.x)$ is at $\alpha$ transition.

@@

There are abstraction for transition variables and application for transition abstraction in \LTP.
They looks like $\Lambda\alpha.M$ and $M\ A$, respectively.
A transition abstraction binds a transition variable in a term.
For example, all $\alpha$ in $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))$ are bound.
It is only natural there is a rule for transition application in \LTP.
The rule is following.

@@

% transition-related symbols disappears when the empty transition is substituted.

@@

% CSP
CSP, cross-stage persistence, is an important feature of multistage calculus.
It enable us to embed value at an outer transition into an inner transition.
\red{inner / outer は適切か?}
$\%$ is dedicated for CSP in \LTP.
For example, $\lambda a:\I.\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x+\%_\alpha a))$

@@

There is another important function called program residualization in \LTP.
It means that a generated code can be dumped into a file.
We can load the dumped file and run it.
The difficulty is arise when program residualization is used with CSP.
Transition variables are classified to two kinds in \LTP in order to deal with this difficulty.

@@

% Kind, Well-formed kind
In addition to ordinary typing rules like simply typed lambda calculus,
there are kinding rules, well-formed kinding rules, term equivalence rules, type equivalence rules, and kind equivalence rules in \LLF.
Kinding rules and well-formed kinding rules are 
introduced in order to prohibit making illegal types such as $\textrm{Vect}\ \textrm{Bool}$.
For a well-formed type $\tau$, $\G \vdash \tau :: K$ means that $\tau$ has a kind $K$ under the environment $\G$ and 
for a well-formed kind $K$, $\G \vdash K$ means that $K$ is a well-formed kind under en environment $\G$.

% Type Equality
Type equality rules are needed because the type equivalence is not obivious unlike simply typed lambda calculus.
For example, $\textrm{Vect}\ 7$ should be equivalent to $\textrm{Vect}\ (3+4)$
but they are not equivalent seemingly. Thus, we must define equivalence rules.
In \LLF, equivalence is expressed with a symbol of $\E$.
$\G \vdash M \E N$ means a term $M$ and a term $N$ are equivalent under the environment $\G$.
$\G \vdash \tau \E \sigma$ means a type $\tau$ and a type $\sigma$ are equivalent under the environment $\G$.
$\G \vdash K \E J$ means a kind $K$ and a kind $J$ are equivalent under the environment $\G$.

@@

Next, we develop \LMD by extending \LTP with \LLF-like dependent types.
From here, we use the word of stage instead of transition 
because we develop a multistage calculus not a multitransition calculus.
\red{stageのほうが言葉としてふさわしいと言いたい}
There are three technical points in the extension from \LTP to \LMD.

@@

First, the way of handling of constants and base types is diffrence between \LMD and \LTP.
We adopt a signature $\Sigma$ to handle constants and base types.
This is because a signature simplify kinding rules relating to type variables.
A signature $\Sigma$ is composed of pairs of a base type and its kind or a constant and its type.
For example, if you want to use boolean, $\Sigma = \B::*, \text{true}:\B, \text{false}:\B$
\red{具体的な導出例を出したほうがよいか? また、なぜsimpleになるのを書くべきか?}

@@

The biggest problem in extension is handling the CSP symbol $\%_\alpha$ in \LMD.
The equvalence simple rule to handle CSP is \QCsp.
\begin{center}
	\infrule{\G\V M \E N : \tau @A}{\G\V\%_\alpha M \E \%_\alpha N : \tau@{A\alpha}}{\QCsp}
\end{center}
However, this rule isn't enough when the parameters of dependent types are cross-staed.
We will discuss on this problem in later section and solve this problem with the new rule \QPercent.
\begin{center}
	\infrule{\G\V M:\tau@{A\alpha} \andalso \G\V M:\tau@A}{\G\V\%_\alpha M \E M : \tau@{A\alpha}}{\QPercent}
\end{center}
\red{この段落でQ-Cspの持つ問題点を指摘すべきか? ただ、ここでmulmatの例を使うと3章と被る。}

>>>>

In \LTP, quote and unquote are written $\TB_\alpha M$ and $\TBL_\alpha M$, respectively.
The type of $\TB_\alpha M$ is $\TW_\alpha \tau$ when $M$ has $\tau$ type.
The type of $\TBL_\alpha M$ is $\tau$ when $M$ has $\TW_\alpha \tau$.
Please notice that $\TBL_\alpha M$ is well-typed, $M$ has a code type from the typing rule of \LTP.
In addition to normal $\beta$-reduction, there is a reduction rule for unquoted quoted code.
\begin{align*}
    \TBL_\alpha (\TB_\alpha M) \longrightarrow M
\end{align*}
It means when a quoted code fragment is unquoted, it returns back to the original code fragment.
This reduction is called $\longrightarrow_\Lambda$ in \LMD.
\red{この文はここに来るべきなのか? LTPの説明ではないが}

@@

A sequence of transition variables is a transition which is called a stage in \LMD.
In \LTP, all judgment has a transition.
Especially, terms without quoting exist at the empty transition which is represented by $\epsilon$.
For example, $(\lambda x:\I.x)\ (1+2)$ is at $\epsilon$ transition and 
$\TB_\alpha (\lambda x:\I.x)$ is at $\alpha$ transition.

@@

There are abstractions for transition variables and applications for transition abstraction in \LTP.
They look like $\Lambda\alpha.M$ and $M\ A$, respectively.
A transition abstraction binds a transition variable in a term.
For example, all $\alpha$ in $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))$ are bound.
It is only natural there is a rule for transition application in \LTP.
The rule is the following.

@@

% transition-related symbols disappear when the empty transition is substituted.

@@

% CSP
CSP, cross-stage persistence, is an important feature of multistage calculus.
It enables us to embed value at an outer transition into an inner transition.
\red{inner / outer は適切か?}
$\%$ is dedicated to CSP in \LTP.
For example, $\lambda a:\I.\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x+\%_\alpha a))$

@@

There is another important function called program residualization in \LTP.
It means that a generated code can be dumped into a file.
We can load the dumped file and run it.
The difficulty arises when program residualization is used with CSP.
Transition variables are classified into two kinds in \LTP in order to deal with this difficulty.

@@

% Kind, Well-formed kind
In addition to ordinary typing rules like simply typed lambda calculus,
there are kinding rules, well-formed kinding rules, term equivalence rules, type equivalence rules, and kind equivalence rules in \LLF.
Kinding rules and well-formed kinding rules are 
introduced in order to prohibit making illegal types such as $\textrm{Vect}\ \textrm{Bool}$.
For a well-formed type $\tau$, $\G \vdash \tau :: K$ means that $\tau$ has a kind $K$ under the environment $\G$ and 
for a well-formed kind $K$, $\G \vdash K$ means that $K$ is a well-formed kind under an environment $\G$.

% Type Equality
Type equality rules are needed because the type equivalence is not obvious unlike simply typed lambda calculus.
For example, $\textrm{Vect}\ 7$ should be equivalent to $\textrm{Vect}\ (3+4)$
but they are not equivalent seemingly. Thus, we must define equivalence rules.
In \LLF, equivalence is expressed with a symbol of $\E$.
$\G \vdash M \E N$ means a term $M$ and a term $N$ are equivalent under the environment $\G$.
$\G \vdash \tau \E \sigma$ means a type $\tau$ and a type $\sigma$ are equivalent under the environment $\G$.
$\G \vdash K \E J$ means a kind $K$ and a kind $J$ are equivalent under the environment $\G$.

@@

Next, we develop \LMD by extending \LTP with \LLF-like dependent types.
From here, we use the word "stage" instead of "transition" 
because we develop a multi-stage calculus, not a multi-transition calculus.
\red{stageのほうが言葉としてふさわしいと言いたい}
There are three technical points in the extension from \LTP to \LMD.

@@

First, the way of handling of constants and base types is the difference between \LMD and \LTP.
We adopt a signature $\Sigma$ to handle constants and base types.
This is because a signature simplifies kinding rules relating to type variables.
A signature $\Sigma$ is composed of pairs of a base type and its kind or a constant and its type.
For example, if you want to use boolean, $\Sigma = \B::*, \text{true}:\B, \text{false}:\B$
\red{具体的な導出例を出したほうがよいか? また、なぜsimpleになるのを書くべきか?}

@@

The biggest problem in extension is handling the CSP symbol $\%_\alpha$ in \LMD.
The equivalence simple rule to handle CSP is \QCsp.
\begin{center}
    \infrule{\G\V M \E N : \tau @A}{\G\V\%_\alpha M \E \%_\alpha N : \tau@{A\alpha}}{\QCsp}
\end{center}
However, this rule isn't enough when the parameters of dependent types are cross-staed.
We will discuss this problem in a later section and solve this problem with the new rule \QPercent.
\begin{center}
    \infrule{\G\V M:\tau@{A\alpha} \andalso \G\V M:\tau@A}{\G\V\%_\alpha M \E M : \tau@{A\alpha}}{\QPercent}
\end{center}
\red{この段落でQ-Cspの持つ問題点を指摘すべきか? ただ、ここでmulmatの例を使うと3章と被る。}
